Monte Carlo tree search compared to A\(^*\) in airspace configuration decision problem

### 2018-05-01 article/article.tex
(+) Completed first draft

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(\mathcal{N}\): the set of nodes;
  \item $h \colon \mathcal{N} \to \mathbb{R}^+$: heuristic estimating the cost
    from a node \(u\) to a final node;
  \item \(\mu_v\): mean value of outcomes of simulations run through or from a
    node \(v\);
  \item \(T_v\): number of simulations run through or from a node \(v\);
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

\subsection{Monte Carlo tree search}
\subsubsection{Overview}
Monte-Carlo is a best-first search method using stochastic simulations. The
algorithm actually uses two trees, an underlying tree associated to the model
(e.g.\ a game tree) and a search tree. The latter is built incrementally by the
algorithm the following way, at each step, starting from the root, the most
promising nodes towards the bottom of the tree are selected. Once a leaf of the
search tree is reached, a new node coming from the model tree is added to the
leaf. A simulation is then run from this newly added node to evaluate the
outcome of a path starting from it. The result of the simulation is eventually
backpropagated to the consecutive parents of the new node.

The above mentioned steps are denoted
\begin{compactenum}
\item selection (or tree walk): choosing successively most promising nodes,
\item expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing successively nodes from the model
  tree,
\item backpropagation (or backup): applying the result of the simulation to
  the search tree.
\end{compactenum}
Those phases are repeated until a stopping criterion (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts} where the tree policy
aggregates phases 1 and 2 to create a new node of the search tree and the
default policy gives an evaluation of the newly added node.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{}\(\pi \gets\) \Call{TreePolicy}{$v_0$}
    \State{}\(v \gets \first(\pi)\)
    \State{}\(\Delta \gets\) \Call{DefaultPolicy}{$v$}
    \State{}\Call{Backup}{$\pi$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose the most promising node among the children
available\footnote{i.e.\ possibly leading to the best evaluation} according to
a \emph{tree policy}. This type of problem can be solved by bandits methods.
Those methods consist in, given a bandit in front of several slot machines
(multi armed bandit), decide which machine will bring the highest reward knowing
the past result. The objective of bandits methods is thus to maximise the reward
and minimise the regret of not playing the best machine.

The bandit problem has been applied to MCTS via the Upper Confidence Tree
algorithm in~\cite{kocsis2006bandit} using the UCB1 equation~\ref{eq:ucb1}.
Let \(u\) be the node from which a child \(v\) must be selected to go deeper in
the tree, \(T_u\) the number of simulations carried out from node \(u\)
(which includes any simulation from nodes in any subtree of \(u\)) and
\(\beta\) a constant. The selected child \(v\) is thus the one maximising the
\emph{UCB1} equation
\begin{equation}
  \label{eq:ucb1}
  \mu_v + 2 \beta \sqrt{\frac{2 \log T_u}{T_v}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficiency in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_v +
  \sqrt{%
    \frac{\beta\log T_u}{T_v}
    \min\left( \frac{1}{4}, \sigma^2_v +
    \sqrt{%
      \frac{2 \log T_u}{T_v}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\subparagraph{Exploration exploitation trade-off} The constant \(\beta\) answers
to the exploration v exploitation dilemma. In the UCB1 equation~\ref{eq:ucb1},
the left hand term increases the value of less explored nodes to consider them
as still promising. A high \(\beta\) value will therefore make the algorithm
prone to try unvisited nodes while a low value will consider almost exclusively
the results obtained so far, even if other paths are better.

\paragraph{Expansion (algorithm~\ref{alg:expansion})}
The above step is carried out only if, considering a node \(v\), a simulation
has been run through each child available from \(v\). In the case where one or
more children haven't been explored yet, one of the unexplored children is
chosen and a simulation is started from it.
\begin{algorithm}
  \caption{Expansion}\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{$v$}
    \State{}$\ell \gets \{ v | v \text{\,children of\,} u, T_v = 0 \}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Simulation and backpropagation}
\paragraph{Simulation}
Once a node has been expanded, random nodes are chosen successively until a
terminal state is found. The cost of the resulting path, from the root to the
node is then evaluated and backpropagated to all the ancestors of the expanded
node.

@@

\subsection{Sequence building}
\subsubsection{Node selection}\label{sssec:node_selection}
The building of the final sequence of nodes differs from the selection on the
aim, the selection chooses the most promising node while the final sequence
must be composed of the best. This means that the final path build won't use the
UCB equation. Chaslot \textit{et al.}~\cite{browne2012survey} propose several
methods to select nodes,
\begin{compactitem}
  \item max-child: select the child with highest reward;
  \item robust child: select the most visited root child;
  \item secure child: select the child maximising a lower confidence bound.
\end{compactitem}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then an mcts is called back with node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_v | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{N}^2 \to \mathbb{R}\). The
function \(\first(G)\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \first(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Theoretical performances}
In this paper, we consider a high branching factor problem.

Let \(M\) be the number of edges between the inital state and a final state and
\(K\) the branching factor.
Then, the complexity of the A\(^*\) is expected to be, in the worst case,
exponential in \(M\) (see~\cite{alliotschiex2002ia&it}).

Since the Monte carlo tree is a stochastic method, one must wait for a
satisfactorily converging solution rather than the exact solution. It has
however been proven in~\cite{kocsis2006bandit} that the bias of the expected
payoff is \(O\left(\frac{\log n}{n}\right)\). Furthermore, the failure
probability\footnote{probability of not selecting the best child} converges to
zero at a polynomial rate.

@@

@@

In the configuration problem, the airspace needs to be devided into sectors. Each
sector represent a area controlled by two officers and is composed of
one or more elementary modules. A partition $P$ is a set of
sectors covering all the airspace.

A sector is a group of elementary modules, but in practice only a collection
of sectors are allowed in this model. This collection is referred as the context.
This restriction reflect the fact the shape of a sector needs have a relatively
'nice' shape in order to be easely manageable for a controller.

\subsubsection{Transitions}

During the day, and with the evolution of the trafic, the workload (rigorously
defined in \ref{sub:workload}) vary. To help controllers to maintain a
homogeneous level of workload, the airspace partitioning needs to be adapt
and therefore make transition from an inital partition to a target partition.
This maneuver requires coordination hence increase controller's workload.
This is the reason why a full reconfiguration isn't doable in practice, and
only a selection of transitions can be operate.

In our model, three transistions are possible namely a merge, a transfer
and a split. For an airspace composed of three modules $A$, $B$, and $C$,
those actions represent:
\begin{itemize}
  \item merge: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A, B, C\}\}$ 
  \item split: $\{\{A, B, C\}\} \quad \rightarrow \quad \{\{A, B\}, \{C\}\}$ 
  \item transfer: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A\}, \{B, C\}\}$ 
\end{itemize}

TODO : meilleur explication des transition

@@

Let us now precise the model established to evaluate a workload metric. The
workload depends on external criteria such as sector and trafic complexity
or humain factors (e.g. controller fatigue, stress). In our model, we assum
that the workload felt by a controller is directly and only related to the trafic
complexity.

The simplest approch considers only the number of aircraft per sector $Nb_{aircraft}$. We
defined two arbitary thresholds, namely $th_{low}$ and $th_{high}$, defining
three zones where the workload is low (resp. normal and high) if $Nb_{aircraft} < th_{low}$
(resp. $th_{low} < Nb_{aircraft} < th_{high}$ and $th_{high} < Nb_{aircraft}$).
Those workload levels are translated to probabilities $p_{low}$, $p_{normal}$ and
$p_{high}$, where $p_{low}(S)$ is the probability that the sector $S$ is underloaded
(resp. balanced and overloaded). In short, for a sector $S$ at time $t$ and $Nb_{aircraft}$:
\begin{equation}
  \begin{cases}
    p_{low}^{S, t} = \mathbb{1}_{[0, th_{low}]}(Nb_{aircraft}) \\
    p_{normal}^{S, t} = \mathbb{1}_{[th_{low}, th_{high}]}(Nb_{aircraft})\\
    p_{high}^{S, t} = \mathbb{1}_{[th_{low}, \infty[}(Nb_{aircraft})
  \end{cases}
\end{equation}
where $\mathbb{1}$ is the indicator function defined by:
\begin{equation}
  \mathbb{1}_{I}(x) = 
  \begin{cases}
    1 & \text{if\,} \quad x \in I\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

This expression of the workload, expressed in term of probabilities, allow us to
use a more complex model for the workload such as a neural network
(see~\cite{gianazza2010forecasting}). This method learn the probabilities $p_{low}$,
$p_{normal}$ and $p_{high}$ based on parameters including trafic complexity and
sector complexity (e.g. airspace volume).

@@

In order to evaluate a given partition $P$ at time $t$, a cost $C(P, t)$ needs
to be defined. This cost depend on the workload in each the sector. We give the
following definitions~\cite{ferrari2017} to represent a high
(respectively normal and low) cost for the partition:
\begin{itemize}
  \item $c_+(P, t) = \sum_{S \in P_t} \delta_{h}(S, t) p_{high}^{S, t} |S|^2$
  \item $c_=(P, t) = 1 / \left(
      \sum_{S \in P_t} \delta_{n}(S, t) p_{norm}^{S} |S|^{-2}
    \right)$
  \item $c_-(P, t) = \sum_{S \in P_t} \delta_{l}(S, t) p_{low}^{S, t} |S|^{-2}$
\end{itemize}
with $\delta_{h}(P, t)$ (resp. $\delta_{n}(P_t)$ and $\delta_{l}(P_t)$) equal 1
if the probability $p_{high}$ (resp. $p_{normal}$ and $p_{low}$) is superior to
the two others, and 0 otherwise.

@@

The parameters $\alpha$, $\beta$, $\gamma$ and $\lambda$ (all positive)
determine a priority on which parameter to optimize. For instance, a high
value $\alpha$ help to
minimize $c_+$, hence the overall number of sectors with too much tafic.
In a real application, it maybe interesting to order those parmaeters as
follow: $\alpha > \gamma > \beta > \lambda$.

@@

In an operational context, each reconfiguration raise the workload for
controllers. This trend needs to be considered and a transition cost needs to
be defined for this matter. A transition is (TODO), and the transition cost is
then:
\begin{equation}
  C_{tr}(P_1, P_2) =
  \begin{cases}
    0 & \text{if\,} \quad P_1 = P_2\\
    1 & \text{otherwise}
  \end{cases}
\end{equation}

@@

With a cost defined for partition and transition, it is now possible to
aggregate everything in order to build a cost function over an entire
path. For a path $\pi = [P_0, \dots, P_n]$ with a time from $t_0$ to $t_n$,
the cost function is given by:
\begin{equation}
  \begin{split}
    f(\pi) = C(P_0, t_0) + \sum_{i = 1}^{n} [C(P_i, t_i) +\\
    \theta C_{tr}(P_{i-1}, P_i)]
  \end{split}
\end{equation}

@@

The Monte Carlo Tree search algorithm maximize a objective function. This function
represent the reward of a path, i.e.\ a list of partitions over time.
Given the loss function defined previously, the target function can be
construct as:
\begin{equation}
  Q(\pi) = \frac{1}{f(\pi)}
\end{equation}

\subsection{Heuristic}
To use effectively the A\(^*\) algorithm, a heuristic is needed. Using the null
heuristic (\(\forall u \in \mathcal{N}, h(u) = 0\)) revealed to be inefficient
considering the branching factor. To build a minimal heuristic, the limitations
introduced in~\ref{ssec:partitions} are not considered, neither those associated
to the previous partition (split, merge or transfer). This allows to generate
the sequence of the best partitions among all possible at each time step.

>>>>

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(C_{tr}(P_1, P_2)\): cost of transition between partitions \(P_1\) and
    \(P_2\);
  \item \(\pi = [P_1, \dots, P_n]\): a sequence of partitions (called \emph{path}
      in the tree search algorithms);
  \item \(f(\pi)\): cost associated to sequence of partitions \(\pi\);
  \item \(\mathcal{N}\): the set of nodes;
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item $h \colon \mathcal{N} \to \mathbb{R}^+$: heuristic estimating the cost
    from a node \(u\) to a final node;
  \item \(\mu_v\): mean value of outcomes of simulations run through or from a
    node \(v\);
  \item \(T_v\): number of simulations run through or from a node \(v\);
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

\subsection{Monte Carlo tree search}
\subsubsection{Overview}\label{sssec:overview}
Monte-Carlo is a best-first search method using stochastic simulations. The
algorithm is based on the computation of the reward expectancy of paths which
is estimated through Monte-Carlo simulations.
The
algorithm actually uses two trees, an underlying tree associated to the model
(e.g.\ a game tree) and a search tree. The latter is built incrementally, each
step being composed of four phases, namely
\begin{compactenum}
\item\label{it:selection} selection (or tree walk): choosing successively
  most promising nodes from the search tree,
\item\label{it:expansion} expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing successively nodes from the model
  tree, from the expanded node to a leaf,
\item backpropagation (or backup): applying the result of the simulation to
  the previously selected nodes of the search tree (phases~\ref{it:selection}
  and~\ref{it:expansion}).
\end{compactenum}
Those phases are repeated until a stopping criterion (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts} where the tree policy
aggregates phases 1 and 2 to create a new node of the search tree and the
default policy gives an evaluation of the newly added node.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{}\(\pi \gets\) \Call{TreePolicy}{$v_0$}
    \State{}\(v \gets \first(\pi)\)
    \State{}\(\Delta \gets\) \Call{DefaultPolicy}{$v$}
    \State{}\Call{Backup}{$\pi$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to build a path from the root of the search tree
by choosing successively the most
promising\footnote{i.e.\ possibly leading to the best evaluation} node.
Given a node \(u\) that has previously been selected, the
best node --
according to a \emph{tree policy} -- among the children of \(u\) is
chosen.
This type
of problem can be solved by bandits methods.
Those methods consist in, given a bandit in front of several slot machines
(multi armed bandit), deciding which machine will bring the highest reward
knowing
the past results. The objective of bandits methods is thus to maximise the reward
and minimise the regret of not playing the best machine.

The bandit problem has been applied to MCTS via the Upper Confidence Tree
algorithm in~\cite{kocsis2006bandit} using the UCB1 equation~\ref{eq:ucb1}.
Let \(u\) be the node from which a child \(v\) must be selected to go deeper in
the tree, \(T_u\) the number of simulations carried out from node \(u\)
(which includes any simulation from nodes in any subtree of \(u\)) and
\(\beta\) a chosen constant. The selected child \(v\) is the one maximising the
\emph{UCB1} equation
\begin{equation}
  \label{eq:ucb1}
  \mu_v + 2 \beta \sqrt{\frac{2 \log T_u}{T_v}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficiency in one player games or sequencing problems. An
alternative using the standard deviation \(\sigma_v\) of the outcome of the
previous simulations involving node \(v\) is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}\label{eq:ucb1-tuned}
  \mu_v +
  \beta\sqrt{%
    \frac{\log T_u}{T_v}
    \min\left( \frac{1}{4}, \sigma^2_v +
    \sqrt{%
      \frac{2 \log T_u}{T_v}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\subparagraph{Exploration exploitation trade-off} The constant \(\beta\) answers
to the exploration-exploitation dilemma. In the UCB1 equations~\ref{eq:ucb1}
and~\ref{eq:ucb1-tuned},
the right hand term increases the UCB value of less explored nodes to consider them
as still promising. A high \(\beta\) value will therefore make the algorithm
prone to try unvisited nodes while a low value will consider almost exclusively
the results obtained so far, even if other paths are better but unexplored.

\paragraph{Expansion (algorithm~\ref{alg:expansion})}
If the selected node has one or more unvisited children, the selection stops and
one of them is added to the search tree randomly, the latter being thus expanded
by this new node.
\begin{algorithm}
  \caption{Expansion}\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{$v$}
    \State{}$\ell \gets \{ v | v \text{\,children of\,} u, T_v = 0 \}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Simulation and backpropagation}
The two steps described in this paragraph aim to guess the reward that can be
expected from a path including a given node.
\paragraph{Simulation}
Once a node has been expanded, random nodes are chosen successively until a
terminal state is found. The cost of the resulting path, from the root to the
node is then evaluated and backpropagated to all the ancestors of the expanded
node.

@@

\subsection{Sequence building}
Once the stopping criterium mentioned in~\ref{sssec:overview} is matched, the
sequence of states can be extracted from the search tree.

\subsubsection{Choosing nodes}\label{sssec:node_selection}
To build the final sequence, nodes are chosen according to a criterion.
Chaslot \textit{et al.} in~\cite{chaslot2008progstrat} propose several
methods to select nodes,
\begin{compactitem}
  \item max-child: select the child with highest mean reward;
  \item robust child: select the most visited root child;
  \item secure child: select the child maximising a lower confidence bound.
\end{compactitem}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then the MCTS algorithm is called back with
node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_v | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{N}^2 \to \mathbb{R}\). The
function \(\first(G)\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \first(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Theoretical performances}
In this paper, a high branching factor problem is considered.

Let \(H\) be the height of the tree, \(K\) the branching factor.
Then, the complexity of the A\(^*\) is expected to be, in the worst case,
exponential in \(H\) (see~\cite{alliotschiex2002ia&it}).

Since the Monte carlo tree is a stochastic method, one must wait for a
satisfactorily converging solution rather than the exact solution. It has
however been proven in~\cite{kocsis2006bandit} that the bias of the expected
payoff tends to zero.

@@

\subsubsection{Time step}
Since the overall goal is to compute a temporal sequence of partitions over a
day, the day has to be sampled. This sampling in represented as the temporal
difference between a node and its children in the trees mentioned in the
previous section.

@@

In the configuration problem, the airspace needs to be divided into sectors. Each
sector represents an area controlled by two controllers and is composed of
one or more elementary modules. A partition $P$ is a set of
sectors covering all the airspace.

A sector is a group of elementary modules, but in practice only a collection
of sectors are allowed in this model. This collection is referred as the context.
This restriction reflects the fact the more complex the shape is, the harder it
is to manage.

\subsubsection{Transitions}\label{sssec:transitions}

During the day, and with the evolution of the trafic, the workload (rigorously
defined in~\ref{sub:workload}) vary. To help controllers to maintain a
homogeneous level of workload, the airspace partitioning needs to be adapted
through a transition from an inital partition to a target partition.
This maneuver requires either coordination or the opening of a new position
and hence increases controller's workload.
This is the reason why a full reconfiguration isn't feasible, and
only a subset of all available transitions can be operated.

In the model considered here, three transistions are possible namely a merge,
a transfer
or a split. For an airspace composed of three modules $A$, $B$, and $C$,
those actions represent:
\begin{itemize}
  \item merge: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A, B, C\}\}$ 
  \item split: $\{\{A, B, C\}\} \quad \rightarrow \quad \{\{A, B\}, \{C\}\}$ 
  \item transfer: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A\}, \{B, C\}\}$ 
\end{itemize}

TODO : meilleure explication des transitions?

Each transition is composed of at most one merge, split or tranfer; more would
result in too complex reconfigurations.

@@

Let us now precise the model established to evaluate a workload metric. The
workload depends on external criteria such as sector and trafic complexity
or humain factors (e.g. controller health or stress). In the model assumed here,
the workload felt by a controller is directly and only related to the trafic
complexity.

The simplest approch considers only the number of aircraft per sector $N_{aircraft}$. We
defined two arbitary thresholds, namely $th_{low}$ and $th_{high}$, defining
three zones where the workload is low (resp. normal and high) if $N_{aircraft} < th_{low}$
(resp. $th_{low} < N_{aircraft} < th_{high}$ and $th_{high} < N_{aircraft}$).
Those workload levels are translated to probabilities $p_{low}$, $p_{normal}$ and
$p_{high}$, where $p_{low}(S)$ is the probability that the sector $S$ is underloaded
(resp. balanced and overloaded). In short, for a sector $S$ at time $t$ and $N_{aircraft}$:
\begin{equation}
  \begin{aligned}
    p_{low}^{S, t} &= \mathbb{1}_{[0, th_{low}]}(N_{aircraft}) \\
    p_{normal}^{S, t} &= \mathbb{1}_{[th_{low}, th_{high}]}(N_{aircraft})\\
    p_{high}^{S, t} &= \mathbb{1}_{[th_{low}, \infty[}(N_{aircraft})
  \end{aligned}
\end{equation}
where $\mathbb{1}$ is the indicator function defined by:
\begin{equation}
  \mathbb{1}_{I}(x) = 
  \begin{cases}
    1 & \text{if\,} \quad x \in I\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

This expression of the workload, expressed in term of probabilities, allow us to
use a more complex model for the workload such as a neural network
(see~\cite{gianazza2010forecasting}). This method learns the probabilities $p_{low}$,
$p_{normal}$ and $p_{high}$ based on parameters including trafic complexity and
the complexity of the sector (e.g. airspace volume).

@@

In order to evaluate a given partition $P$ at time $t$, a cost $C(P, t)$ needs
to be defined. In this paper, this cost depends on the workload in each the
sector. The definitions given in~\cite{ferrari2017} are used to represent a high
(respectively normal and low) cost for partition \(P\):
\begin{equation}
  \begin{aligned}
    c_+(P, t) &= \sum_{S \in P_t} \delta_{h}(S, t) \cdot p_{high}^{S, t} \cdot |S|^2\\
    c_=(P, t) &= \left(
    \sum_{S \in P_t} \delta_{n}(S, t) \cdot p_{norm}^{S} \cdot |S|^{-2}
    \right)^{-1}\\
    c_-(P, t) &= \sum_{S \in P_t} \delta_{l}(S, t) \cdot p_{low}^{S, t} \cdot |S|^{-2}
  \end{aligned}
\end{equation}
with $\delta_{h}(P, t)$ (resp. $\delta_{n}(P_t)$ and $\delta_{l}(P_t)$) equals 1
if the probability $p_{high}$ (resp. $p_{normal}$ and $p_{low}$) is superior to
the two others, and 0 otherwise.

@@

The parameters $\alpha$, $\beta$, $\gamma$ and $\lambda$ (all positive)
determine a priority on which parameter to optimize. For instance, a high
value $\alpha$ help to
minimize $c_+$, hence the overall number of sectors with too much trafic.
In a real application, it maybe interesting to order those parameters as
follow: $\alpha > \gamma > \beta > \lambda$.

@@

In an operational context, each reconfiguration increases the workload for
controllers. This needs to be considered via the definition of a transition
cost.
A transition is the difference of two partitions separated by only on time step.
The transition cost is then
\begin{equation}
  C_{tr}(P_1, P_2) =
  \begin{cases}
    0 & \text{if} \quad P_1 = P_2\\
    1 & \text{otherwise}
  \end{cases}
\end{equation}

@@

Having defined a cost for partitions and transitions, it is now possible to
aggregate everything in order to build a cost function over an entire
path. For a path $\pi = [P_0, \dots, P_n]$ with a time from $t_0$ to $t_n$,
the cost function is given by:
\begin{equation}
  \begin{split}
    f(\pi) = C(P_0, t_0) + \sum_{i = 1}^{n} [C(P_i, t_i) +\\
    \theta \cdot C_{tr}(P_{i-1}, P_i)]
  \end{split}
\end{equation}

@@

The Monte Carlo Tree search algorithm maximizes an objective function. This function
can be interpreted as the reward of a path.
Given the loss function defined previously, the target function can be
constructed as:
\begin{equation}
  Q(\pi) = \frac{1}{f(\pi)}
\end{equation}

\subsection{Heuristic}
To use effectively the A\(^*\) algorithm, a heuristic is needed. Using the null
heuristic (\(\forall u \in \mathcal{N}, h(u) = 0\)) revealed to be inefficient
considering the branching factor. To build a minimal heuristic, the limitations
introduced in~\ref{sssec:transitions} are not considered, neither those associated
to the previous partition (split, merge or transfer). This allows to generate
the sequence of the best partitions among all possible at each time step.

\section{Results}

TODO
